\documentclass[class=article,crop=false]{standalone} 
\input{../preamble.tex}

\begin{document}
\subsubsection{Dual of linear programs}
The primal (P) is:
\begin{align*}
\max_{x \in \rr^{n}}\quad & \langle c,x \rangle \\
\text{subject to } \quad &x \geq 0\\
y:\quad  &Ax \leq b
\end{align*}
The dual (D) is:
\begin{align*}
\min\quad &\langle b,y \rangle \\
\text{subject to } \quad & y\geq 0\\
x:\quad & A^{T} y \geq c
\end{align*}

Rules to transform (P) to (D):
\begin{enumerate}[label=(\arabic*)]
	\item $ \max \to \min$ and vice versa.
	\item variables $ \to$ constraints and vice versa.
	\item objective and RHS of inequality flip places.
	\item matrices transpose.
\end{enumerate}
"SOB" mnemonic: sensible, odd, and bizarre from a business perspective
\begin{align*}
\text{ primal variable } x_i \quad &x_i\geq 0 \text{: sensible}  \\
& \text{ no constraint: odd}   \\
&x_i \leq 0 \text{: bizarre} 
\end{align*}
\begin{align*}
	\text{ constraints in primal when maximizing} \quad & a_i^{T}x\leq b_i \text{: sensible (think budget)}  \\
& a_i^{T}x=b_i\text{: odd}   \\
&a_i^{T}x \geq b_i \text{: bizarre} 
\end{align*}
The rule: a dual constraint is S/O/B if primal variable is S/O/B. And vice versa. 

\begin{eg}
\begin{align*}
	(P) \qquad \qquad \min_{\substack{x\geq 0\\ x \in \rr^{2}}} \quad &3x_1+2x_2\\
						 y_1: \quad &x_1 + 2x_2 \geq 5 \qquad S\\
						 y_2:\quad & \underbrace{x_2 \leq 2 \qquad}\qquad B\\
							   & \begin{pmatrix} 1&2\\0&1 \end{pmatrix} \begin{pmatrix} x_1\\x_2 \end{pmatrix} 
\end{align*}
\begin{align*}
	(D) \qquad \qquad \min_{\substack{y \in \rr^{2}\\y_1\geq 0\ S\\ y_2 \leq 0 \ B}} \quad &5y_1+2y_2\\
						 x_1: \quad &y_1 + 0y_2 \leq 3 \qquad S\\
						 x_2:\quad & 2y_1+y_2 \leq 2 \qquad  S
\end{align*}


Observe: $ f(x) = 3x_1 + 2 x_2 = \underbrace{2x_1}_{\geq 0} + \underbrace{(x_1 + 2x_2)}_{\geq 5} \geq 5$. We proved $ 5\leq p^* $. However, this is not the tightest bound. The dual variables give us the tightest: 3 times the first constraint and -4 times the second constraint yields $ 7\leq p^* $.
\end{eg}
\allbold{Duality gap}: $ x, \lambda$ feasible, $ f_0(x) - g(\lambda,\nu)$.

\subsubsection{Strong duality results}
\begin{itemize}
	\item If $ (P)$ isn't convex, strong duality is unlikely except certain nonconvex QP: s-lemma/s-procedure (see Appendix of BV).
	\item  If (P) is convex, strong duality holds under certain constraint qualifications (CQ) such as Slater's condition.
		\begin{align*}
		\min\quad &f_0(x) \\
		\text{subject to } \quad &f_i(x) \leq 0, i = 1,\ldots,m \\
		&Ax = b
		\end{align*}
		\begin{defn}[Slater's conditions]
			They hold if there exists a strictly feasible point, $ x \in \ri(\dom(f_0))$ and

			if $ f_i$ is affine, $ f_i(x)\leq 0$ (feasible)

			if $ f_i$ isn't affine, $ f_i(x)<0$ (strictly feasible)

			and $ Ax=b$.
		\end{defn}
		\begin{thm}
			If (P) is convex and Slater's conditions hold, then
			\begin{enumerate}[label=(\roman*)]
				\item we have strong duality, $ d^* =p^* <\infty$
				\item there exists an optimal solution to the dual problem.
			\end{enumerate}
		\end{thm}
		\begin{note}
			Slater's does NOT imply there exists an optimal \emph{primal} solution.
		\begin{eg}
		$ \inf_{x \in \rr} e^{x}$. It is convex, lsc, proper. But it is not coercive so it doesn't have an optimal primal solution.
		\end{eg}
		\end{note}
		\begin{remark}
		Often we want Slater's condition on the dual. Since the dual of the dual is the primal, then we guarantee an optimal solution.
		\end{remark}
		\begin{coro}[Slater for LP]
		Slater's conditions hold iff the LP is feasible \emph{i.e.} $ p^* < \infty$.

		$ p^* < \infty \implies d^* =p^* $ and dual optimal solution exists.

		$ d^* >- \infty \implies d^* =p^* $ and primal optimal solution exists.

		Hence if either $ p^* $ or $ d^* \in \rr$ (not $ \pm \infty$), then optimal primal and dual solutions exist.
		\end{coro}
		\begin{note}
		$ d^* = - \infty, p^* = -\infty$ is possible but rare. This is not strong duality.
		\end{note}
\end{itemize}
\end{document}
