\documentclass[class=article,crop=false]{standalone} 
\input{../preamble.tex}

\begin{document}
\subsection{by hand}
\subsubsection{implicit differentiation}

$ F(x,y) = 0$ often implicitly defines a function  $ y = f(x)$. Find  $ dy /dx$.

 $ F = 0 \implies dF / dx =  d0 / dx =0$. Thus,
 \begin{align*}
 	0=\frac{d F}{d x} &= \frac{\partial F}{\partial x} \frac{d x}{d x} + \frac{\partial F}{\partial y} \frac{d y}{d x}  \\
	0&= \frac{\partial F}{\partial x} + \frac{\partial F}{\partial y} \frac{dy}{dx} \\
	\frac{d y}{d x} &= - \frac{ \frac{\partial F}{\partial x} }{ \frac{\partial F}{\partial y} }
 \end{align*}

\subsubsection{matrix variables}
See BV04, or A Matrix Handbook for Statisticians Seber 08, or Matrix Cookbook, for examples.
\begin{eg}
\begin{align*}
	\nabla (\log \det (X)) = X^{-1}, X \succ 0
\end{align*}
\end{eg}
\subsubsection{parametric functions}
See notes for details.
\begin{align*}
	f(x) = \max_{z \in Z} \phi(x,z)
\end{align*}
\begin{thm}[Danskin]
	Suppose $ Z$ compact,  $ \phi$ jointly continuous and $ \phi( \cdot ,z)$ convex. Define
	\begin{align*}
		Z(x) = \argmax_{z \in Z} \phi(x,z).
	\end{align*}
	Then
	\begin{enumerate}[label=(\arabic*)]
		\item The directional derivative $ D_d$ satisfies
			 \begin{align*}
				 D_d f(x) = \max_{z \in Z(x)} D_d \phi(x,z)
			\end{align*}
		\item If $ \phi( \cdot ,z)$ is differentiable, $ \nabla _x \phi$ is continuous, then
			\begin{align*}
				\partial f(x) = \conv \{\nabla _x \phi(x,z): z \in Z(x)\} 
			\end{align*}
			and $ Z(x)$ a singleton  $ \implies$ $ f$ is differentiable.
	\end{enumerate}
\end{thm}
\begin{note}
This theorem doesn't apply to the discrete case.
\end{note}
\begin{thm}[Dubovitskii and Milyutin]
	If $ |Z|$ is finite,  $ \phi( \cdot ,z)$ is convex $ \ \forall \ z \in Z$, then
	\begin{align*}
		\partial f = \conv \left\{ \bigcup_{ z \in Z(x)} \partial \phi(x,z) \right\} 
	\end{align*}
\end{thm}
\begin{eg}
	$ f(x) = |x| = \max \{x,-x\} $. Then $ \partial f(0) = \conv \{1,-1\} = [-1,1] $.
\end{eg}

\begin{align*}
	f(x) = \min_{z \in Z} \phi(x,z)
\end{align*}
\begin{thm}
Under the same conditions,
\begin{align*}
	\partial f(x) = \partial \phi(x,z) \ \forall \ z \in Z(x)
\end{align*}
\end{thm}

\begin{align*}
	f(x) = \int \phi(x,z) dz
\end{align*}
\begin{thm}
Sometimes,
\begin{align*}
	f'(x) = \int \frac{d}{dx} \phi(x,z) dz 
\end{align*}
\end{thm}
\subsection{Adjoint state method}
\begin{enumerate}[label=(\arabic*)]
	\item implicit differentiation and careful parentheses.
		\begin{align*}
			\min_{p} g(u(p),p)\ s.t.\ f(u(p),p) = 0, u \in \mathcal{ H}
		\end{align*}
		\begin{eg}
			$ u_{tt} = c^2(x) u_{x x}$. Let $p= c^2(x)$, \emph{e.g.} varying speed of sound. Or $ p$ could be other parameters like ICs and BCs. Applications in inverses problems such as oil detection. This is often called ``PDE-constrained optimization problem''.
			To solve this, we ignore the constraint first. Then we wish to solve $ dg /dp$.
		\end{eg}
		\begin{eg}
		\begin{align*}
			\min g(u,p) &\ s.t.\ A u = b, A= V \text{ diag}  (p) V^{T}, p \in \rr^{n}, u \in \rr^{m}, A \in \rr^{m} \times \rr^{m}\\
			g(u,p) &= \frac{1}{2} \sum_i (u_i - y_i)^2 + \frac{1}{2} \norm{ p}^2 \\
		\end{align*}
		\end{eg}
		\begin{eg}
			Let $ \mathcal{ H} = \rr^{m}$. $ f(u,p) = A(p) \cdot  u -b(p)=0$, linear in $ u$, so $ u(p) = A(p)^{-1} b(p)$. The goal is to find
			\begin{align*}
				(\nabla _p g)^{T} &= g_p + g_u u_p \\
				0=f_p &= A_p u + Au_p -b_p \\
				u_p &= A^{-1}(b_p - A_p u)\\
				(\nabla _p g)^{T} &= g_p + g_u (A^{-1}(b_p - A_p u))\\
						  &= g_p + (g_u A^{-1})(b_p - A_p u) \\
			\end{align*}
			Let $ \lambda^* = g_u A^{-1} \implies \lambda = A^{-*} g_u^* $, then we have the adjoint-state equation
			\begin{align*}
				A^* \lambda = g_u^* 
			\end{align*}
			Now by clever grouping we are only solving one RHS instead of $ n$ RHS.
		\end{eg}
	\item adjoints of (bounded/unbounded) linear operators.
		What about adjoints if $ | \mathcal{ H}| = \infty$?
		\begin{eg}
			Let $ L : \mathcal{ H} \to \mathcal{ H}$, $ \mathcal{ H} = L^2[0,T]$:
		\begin{align*}
			L(u) = 3u'+4u,u(0)=0, t \in [0,T].
		\end{align*}
		Let $ f(u,p) = L(u) -h(t)$. To get the formal adjoint (adjoint doesnt exists since $ L$ doesn't have full domain),
		 \begin{align*}
			 \int_0^T (3u'+4u) v dt &= 3\int u'vdt + 4\int uv dt\\
			 \int u(-3v'+ 4v) dt&= 3uv\big|_0^T - 3\int uv' dt + 4 \int uv dt && \text{ set }  v(T)=0
		\end{align*}
		Thus,
		\begin{align*}
			L^* (v) = -3v'+4v, v(T)=0, L^* (v) = h
		\end{align*}
		\end{eg}
		\begin{remark}
		~\begin{enumerate}[label=(\arabic*)]
			\item doesn't require linear PDE's
			\item not always a good idea: memory issues, consistency from the order of applying discretization or optimization
			\item software: Dolfin-Adjoint, FEnICS
		\end{enumerate}
		\end{remark}
\end{enumerate}
\end{document}
